---
title: "pew.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "lumen"
highlight-style: "atom-one"
bibliography: '../assets/epsy8252.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


The data in [pew.csv](https://raw.githubusercontent.com/zief0002/benevolent-anteater/main/data/pew.csv) come from a telephone survey conducted by [The Pew Research Center for The People &amp; The Press](https://www.pewresearch.org/politics/2007/04/15/public-knowledge-of-current-affairs-little-changed-by-news-and-information-revolutions/) in February 2007. The data represent a probability sample of 1,502 adults in the U.S. The attributes in the dataset include:

- `id`: The respondent ID number
- `knowledge`: Score on the *Political News Knowledge Test*; on a scale from 0--100, with higher values indicating more political knowledge.
- `news`: Respondent's level of news exposure; on a scale from 0--100, with higher values indicating more news exposure
- `age`: Respondent's age
- `educ`: Respondent's education level; indicates highest grade completed from 8 to 18 (a post masters degree)
- `male`: Dummy coded sex variable (0 = Not male, 1 = Male)
- `ideology`: Respondent's political ideology; on a scale from 0--100, where 0 is as liberal as possible and 100 is as conservative as possible
- `party`: Respondent's party affiliation
- `state_blue`: Does the respondent live in a "blue state" (voted Democrat) or a "red state" (voted Republican) as classified by voting in the 2004 presidential election
- `engagement`: Index of individual political engagement; on a scale from 0--100, with higher values indicating more political engagement


### Preview

```{r}
# Import data
pew = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/benevolent-anteater/main/data/pew.csv")

# View data
pew
```


### References


