---
title: "same-sex-marriage.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
highlight-style: "atom-one"
bibliography: '../assets/epsy8252.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The data in [same-sex-marriage.csv](https://raw.githubusercontent.com/zief0002/benevolent-anteater/main/data/same-sex-marriage.csv) were collected from a poll taken by Pew Research Center in 2015. These particular data consist of 500 American's responses. The attributes are:

- `support`: Dummy-coded variable indicating whether the respondent supports same sex marriage? (1=Yes; 0=No)
- `friends`: Number of respondent's friends that are gay or lesbian
  + 0 = None
  + 1 = Only 1 or 2
  + 2 = Some
  + 3 = A lot
- `education`: Respondent's level of education
  + 0 = High school or less
  + 1 = Some college (no degree) or two year associate degree from a college or university
  + 2 = Four year college or university degree/Bachelor's degree (e.g., BS, BA, AB)
  + 3 = Some postgraduate or professional schooling or postgraduate or professional degree
- `female`: Dummy-coded variable indicating whether the respondent is female (1=Yes; 0=No)
- `age`: Respondent's age, in years
- `attend`: Frequency the respondent attends religious services
  + 0 = Never
  + 1 = Seldom
  + 2 = A few times a year; 
  + 3 = Once or twice a month; 
  + 4 = Once a week
  + 5 = More than once a week
- `ideology`: Respondent's description of their political views 
  + 1 = Very conservative
  + 2 = Conservative
  + 3 = Moderate
  + 4 = Liberal
  + 5 = Very liberal
- `party`: Respondent's self-identified political party affiliation (Democrat, Republican, Independent)






### Preview

```{r}
# Import data
same_sex = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/benevolent-anteater/main/data/same-sex-marriage.csv")

# View data
same_sex
```


<!-- ### References -->
