---
title: "usnwr-2024.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "lumen"
highlight-style: "atom-one"
bibliography: '../assets/epsy8252.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The data in [usnwr-2024.csv](https://raw.githubusercontent.com/zief0002/benevolent-anteater/main/data/usnwr-2024.csv) includes 13 attributes collected from the n=111 graduate schools of education ranked in the U.S. News and World Report’s 2023--2024 Best Education Graduate Schools.


<br />


## Attributes

- `rank`: Rank in USNWR
- `school`: Graduate program of Education
- `peer_rating`: Peer assessment score from deans', program directors', and senior faculty's judgement of the academic quality of the program on a scale of 1 (marginal) to 5 (outstanding).
- `enroll`: Total enrollment of part-time and full-time graduate students in 2022--23
- `ft_students`: Total number of full-time graduate students enrolled in 2022--23
- `ft_faculty`: Total number of full-time faculty employed in 2022--23
- `nonres_tuition`: Amount of yearly non-resident graduate tuition
- `sf_ratio`: Adjusted ratio of graduate students to faculty in 2022--2023
- `pct_doc`: Percentage of students enrolled in a Ph.D./Ed.D. in 2022--23
- `ug_gpa`: Average undergraduate GPA for Ph.D. students admitted in 2022--23
- `gre`: Averge total GRE score (based on verbal and quantitative) for Ph.D. students admitted in 2022--23
- `doc_accept`: Acceptance rate for doctoral students in 2016
- `tot_pubs`: Total number of publication attributed to faculty in the program from 2019--23 based on SCOPUS
- `tot_res`: Total amount of funded research (in millions of dollars)


<br />


## Preview

```{r}
# Import data
usnwr = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/benevolent-anteater/main/data/usnwr-2024.csv")

# View data
usnwr |>
  print(width = Inf)
```





